---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author: "Group 05: [Bùi Trọng Thành - 24125043, Phan Anh Khoa - 24125059, Văn Tuấn Khải - 24125032, Huỳnh Hữu Phúc Minh - 24125068]"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
params:
  group_number: 05
fontsize: 11pt
geometry: margin=1in
---


# Authorship and contribution statement {-}

We confirm that every group member can explain the submitted analysis. AI use is reported in `GroupXX_AI_Log.csv`, and the group checked every AI-assisted item used in this report.

| Student | ID | Main contributions | Verification performed |
|---|---|---|---|
| [Name] | [ID] | [Work] | [Check] |

# Abstract {-}

*Maximum 150 words: prediction problem, methods, main evidence, and recommendation.*

# Reproducible setup {-}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)

group_number <- as.integer(params$group_number)
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```

## Shared helper functions {-}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration, eval=FALSE}
config <- if (group_number %% 2L == 1L) {
  list(
    file = "fat.csv",
    response = "brozek",
    excluded = c("brozek", "siri", "density", "free")
  )
} else {
  list(
    file = "prostate.csv",
    response = "lpsa",
    excluded = "lpsa"
  )
}

data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```

*Define the prediction task and justify every exclusion.*
Target (Response): The prediction target is brozek.  

Candidate Predictors: The available predictors are the remaining non-excluded anatomical measurements (e.g., age, weight, height, adipos, neck, chest, abdom, hip, thigh, knee, ankle, biceps, forearm, and wrist).

Unit of Observation: A single male subject. 

Study Population: The population of men evaluated in the body-fat study described by Johnson (1996).  

Prediction Objective: To predict an individual's body fat percentage (brozek) relying solely on simple, non-invasive anatomical measurements.

Dataset Source: The original study described by Johnson (1996).  

Required Exclusions & Leakage Justification: We are required to explicitly exclude siri, density, and free. These specific variables encode alternative body-fat calculations or are physical quantities derived using body-fat information.

Including any outcome-derived variable fundamentally invalidates the prediction task by introducing target leakage. If included, the regularized models would not learn a generalized statistical relationship from anatomical features; instead, they would merely reverse-engineer a deterministic algebraic formula, rendering any comparison of OLS, ridge, and lasso meaningless.

## 1B. Split and train-only preprocessing

```{r split-and-scale, eval=FALSE}

if (anyNA(analysis_data)) {
  stop("Missing values detected.")
}

split <- split_rows(nrow(analysis_data), train_prop = 0.80, seed = seeds$split)
train_id <- split$train
test_id  <- split$test

# Isolate raw matrices
x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])

# Apply training scalars to both partitions
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test  <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)

# Isolate responses
y_train <- y_raw[train_id]
y_test  <- y_raw[test_id] # REMAIN LOCKED UNTIL PROBLEM 4D

# Shared 5-fold cross-validation allocation
foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

*Report seeds, dimensions, scaling, folds, and leakage safeguards.*
Data Split: We applied an $80/20$ train/test split utilizing the group-specific seed $240205$. This yields a training sample of $n = 201$ observations and a locked holdout sample of $n = 51$ observations.  

Missing-Value Plan: The dataset contains zero missing values. Had missings been detected, any imputation parameters (such as the median or mode) would be calculated exclusively on the $201$ training observations to prevent holdout distribution data from influencing model inputs.  

Scaling Rule: Regularization imposes penalties ($L_1, L_2$) on coefficient magnitudes. To ensure all candidate predictors face equitable shrinkage, we standardize the design matrix. The transformation is computed as  $z_{i,j} = \frac{x_{i,j} - \mu_{train,j}}{\sigma_{train,j}}$. Crucially, $\mu$ and $\sigma$ are derived exclusively from the training matrix. Applying training constants to the test matrix prevents structural target leakage, ensuring no variance or distributional shifts from the holdout set contaminate the model's feature space.  

Cross-Validation Assignment: We defined a single, reproducible 5-fold assignment (foldid) on the training observations using the seed $240305$. Reusing this exact vector across OLS, Ridge, and Lasso guarantees that models are compared on identical out-of-fold samples, removing variance due to random resampling from the model comparison.  

Leakage Prevention Summary: By sequestering y_test, calculating scaling factors solely on $X_{train}$, and locking a shared cross-validation grid, the model acts entirely blind to the holdout data. This isolates the true generalization error for our final evaluation. 

## 1C. Training-data exploration

```{r training-eda, echo=TRUE, message=FALSE, warning=FALSE}
# Isolate training data to strictly prevent leakage
train_data <- analysis_data[train_id, ]

# 1. Response Distribution
p_dist <- ggplot(train_data, aes(x = brozek)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(title = "Distribution of Brozek Body Fat % (Training Set)",
       x = "Brozek (%)", y = "Count")
print(p_dist)

# 2. Focused Relationship / Correlation Table
# Identify top features correlated with the target
cor_matrix <- cor(train_data)
cor_with_target <- sort(abs(cor_matrix[, "brozek"]), decreasing = TRUE)
knitr::kable(head(cor_with_target, 6), 
             col.names = c("Absolute Correlation with Brozek"), digits = 3,
             caption = "Top Correlated Predictors (Training Set)")

# 3. Anomaly Detection
# Scatter plot to identify severe outliers
p_anom <- ggplot(train_data, aes(x = height, y = weight)) +
  geom_point(color = "darkred", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Anomaly Check: Height vs. Weight",
       x = "Height (inches)", y = "Weight (lbs)")
print(p_anom)

# Isolate the mathematically anomalous observation
anomaly_record <- train_data |> filter(height < 40 | weight > 300)
knitr::kable(anomaly_record[, c("brozek", "age", "weight", "height")], 
             caption = "Potential Influential / Anomalous Observations")

# 4. Multicollinearity Detection
# Find highly collinear predictor pairs (r > 0.85) among independent variables
pred_cor <- cor(train_data |> select(-brozek))
pred_cor[lower.tri(pred_cor, diag = TRUE)] <- NA
high_cor <- as.data.frame(as.table(pred_cor)) |>
  filter(abs(Freq) > 0.85) |>
  arrange(desc(abs(Freq)))

knitr::kable(head(high_cor, 5), 
             col.names = c("Predictor 1", "Predictor 2", "Correlation"), digits = 3,
             caption = "Highly Collinear Predictor Pairs (r > 0.85)")
```

*Interpret multicollinearity, one anomaly, and the question regularization will address.*

Response distribution: The brozek target variable is approximately normally distributed and centered near 19%, indicating that ordinary least squares (OLS) and its regularized variants will not require immediate non-linear transformations to achieve stable residuals.

Focused relationship: Torso measurements demonstrate the strongest linear relationship with the target within the training fold, specifically abdom ($r = 0.817$) and adipos ($r = 0.744$).

Anomalous observation: A mathematically severe anomaly exists (height = 29.5 inches, weight = 205.0 lbs), which is almost certainly a data-entry error. Rather than dropping it, this high-leverage point is retained to empirically evaluate how $L_1$ and $L_2$ penalties stabilize estimation when the design matrix contains structural errors compared to the unpenalized OLS baseline.  

Multicollinearity: Severe pairwise correlations are present among spatial measurements, notably weight and hip ($r = 0.940$), and adipos and abdom ($r = 0.932$). In an OLS model, this collinearity inflates the variance of coefficient estimates and drives up the condition number $\kappa_2(G)$ of the Gram matrix $G = \frac{1}{n}X^{\top}X$.

Concluding Question:
Given the severe multicollinearity among body circumferences, can an $L_1$ penalty (lasso) isolate a sparse, interpretable subset of features without sacrificing the predictive stability provided by an $L_2$ penalty (ridge) that smoothly shrinks highly correlated predictors?

# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

```{r ols, eval=FALSE}
# 1. Fit OLS Model
df_train <- data.frame(y = y_train, x_train)
ols_fit <- lm(y ~ ., data = df_train)

# 2. Extract and Report Coefficients (Idiomatic tidyverse/broom)
ols_coef_table <- broom::tidy(ols_fit)

# Output table for the report
knitr::kable(
  ols_coef_table, 
  digits = 4, 
  caption = "OLS Baseline Coefficients"
)

# 3. Calculate Errors (MSE)
ols_train_mse <- mean(resid(ols_fit)^2)

ols_cv_mse <- purrr::map_dbl(1:5, function(k) {
  val_mask <- (foldid == k)
  
  # Strict isolation: train only on out-of-fold data
  fit_k <- lm(y ~ ., data = df_train[!val_mask, ])
  preds_k <- predict(fit_k, newdata = df_train[val_mask, ])
  
  # Calculate validation MSE for fold k
  mean((df_train$y[val_mask] - preds_k)^2)
}) %>% mean()

# 4. Report Errors
cat(sprintf("Training MSE: %.4f\n", ols_train_mse))
cat(sprintf("5-Fold CV MSE: %.4f\n", ols_cv_mse))

# 5. Diagnostic Plot
plot(ols_fit, which = 1, main = "OLS Residuals vs Fitted")
```

Interpretation of OLS Baseline:
The OLS model exhibits severe instability due to multicollinearity among the physical measurements. Evidence of this ill-conditioning includes inflated standard errors (e.g., weight $SE \approx 1.638$) and physiologically counterintuitive coefficient signs (e.g., a negative coefficient for weight). Because the highly correlated predictors cause the eigenvalues of the Gram matrix to approach zero, the variance of the unpenalized $\hat{\beta}$ estimates explodes. We retain all features as instructed; however, this instability necessitates the $L_1$ and $L_2$ regularization approaches explored in subsequent sections.

## 2B. Ridge

```{r ridge, eval=FALSE}
# Fit Ridge CV using shared folds to prevent data leakage
ridge_cv <- fit_cv_glmnet(x_train, y_train, alpha = 0, foldid = foldid)

# Extract lambdas
ridge_lambda_min <- ridge_cv$lambda.min
ridge_lambda_1se <- ridge_cv$lambda.1se

# Extract CV errors
ridge_err_min <- ridge_cv$cvm[ridge_cv$lambda == ridge_lambda_min]
ridge_err_1se <- ridge_cv$cvm[ridge_cv$lambda == ridge_lambda_1se]

# Bind coefficients for clean reporting
ridge_coefs <- tibble::tibble(
  term = rownames(coef(ridge_cv)),
  lambda_min = as.vector(coef(ridge_cv, s = "lambda.min")),
  lambda_1se = as.vector(coef(ridge_cv, s = "lambda.1se"))
)

# Output formatting for the report
cat(sprintf("lambda.min: %.4f (CV MSE: %.4f)\n", ridge_lambda_min, ridge_err_min))
cat(sprintf("lambda.1se: %.4f (CV MSE: %.4f)\n", ridge_lambda_1se, ridge_err_1se))

knitr::kable(
  ridge_coefs,
  digits = 4,
  caption = "Ridge Coefficients at lambda.min and lambda.1se"
)

# Visualizations: CV Curve and Coefficient Paths
par(mfrow = c(1, 2))
plot(ridge_cv)
title("Ridge CV Error", line = 2.5)

plot(ridge_cv$glmnet.fit, xvar = "lambda", label = TRUE)
abline(v = log(c(ridge_lambda_min, ridge_lambda_1se)), lty = 2, col = c("blue", "red"))
title("Ridge Coefficient Paths", line = 2.5)
par(mfrow = c(1, 1))
```

Interpretation of Ridge Regression Results:

*Tuning and Validation Error*:
Using 5-fold cross-validation on the locked training data, the ridge regression model identified an optimal penalty of lambda.min (CV MSE: 19.7095). Applying the stricter one-standard-error rule yields lambda.1se (CV MSE: 22.1846), which applies heavier regularization in exchange for a slightly higher training estimation error.

*Behavior of Correlated Predictors ($L_2$ Mechanism)*:
The $L_2$ penalty in the ridge objective function, defined as $\frac{\lambda}{2} \vert{}\vert{}b\vert{}\vert{}_2^2$, heavily penalizes large individual coefficients. Consequently, when presented with highly correlated physiological predictors (e.g., weight, abdom, chest), ridge regression exhibits a "grouping effect." Rather than assigning a massive weight to one predictor and zero to another, the $L_2$ norm mathematically prefers to distribute the coefficients somewhat evenly among correlated features, as $a^2 + a^2 < (2a)^2 + 0^2$.
This stabilization is highly visible in the coefficient shifts from lambda.min to lambda.1se. For instance, the heavily inflated abdom coefficient shrinks from 5.0971 to 2.5770. Furthermore, the sign-flipping observed in the ill-conditioned OLS baseline is corrected; the coefficient for weight transitions from a biologically counterintuitive negative value (-0.2324) at lambda.min to a logical positive value (0.3160) at the more heavily penalized lambda.1se.

*Lack of Exact Sparsity*:
It is critical to note that while the $L_2$ penalty shrinks coefficients toward zero asymptotically, it lacks a coordinatewise soft-thresholding mechanism. Therefore, ridge regression does not perform variable selection. Numerically tiny coefficients, such as knee (-0.0032 at lambda.min), remain mathematically active in the prediction equation and must not be interpreted as exact zeros.

## 2C. Lasso

```{r lasso, eval=FALSE}
# 1. Fit Lasso CV using shared folds
lasso_cv <- fit_cv_glmnet(x = x_train, y = y_train, alpha = 1, foldid = foldid)

# 2. Extract optimal lambdas and their CV errors
lasso_lambda_min <- lasso_cv$lambda.min
lasso_lambda_1se <- lasso_cv$lambda.1se

lasso_err_min <- lasso_cv$cvm[lasso_cv$lambda == lasso_lambda_min]
lasso_err_1se <- lasso_cv$cvm[lasso_cv$lambda == lasso_lambda_1se]

# 3. Custom extraction of exact nonzero feature names
get_active_features <- function(fit, s) {
  # Extract sparse matrix of coefficients
  b <- as.matrix(coef(fit, s = s))
  # Subset to non-zero, non-intercept terms
  active <- rownames(b)[b[, 1] != 0 & rownames(b) != "(Intercept)"]
  return(active)
}

lasso_active_min <- get_active_features(lasso_cv, "lambda.min")
lasso_active_1se <- get_active_features(lasso_cv, "lambda.1se")

# 4. Formatted output for the RMarkdown report
cat(sprintf("lambda.min: %.4f (CV MSE: %.4f)\n", lasso_lambda_min, lasso_err_min))
cat("Active predictors (lambda.min):", paste(lasso_active_min, collapse = ", "), "\n\n")

cat(sprintf("lambda.1se: %.4f (CV MSE: %.4f)\n", lasso_lambda_1se, lasso_err_1se))
cat("Active predictors (lambda.1se):", paste(lasso_active_1se, collapse = ", "), "\n")

# 5. Visualizations
par(mfrow = c(1, 2))

plot(lasso_cv)
title("Lasso CV Error", line = 2.5)

plot(lasso_cv$glmnet.fit, xvar = "lambda", label = TRUE)
abline(v = log(c(lasso_lambda_min, lasso_lambda_1se)), lty = 2, col = c("blue", "red"))
title("Lasso Coefficient Paths", line = 2.5)

par(mfrow = c(1, 1))

# Extract coefficients as vectors
lasso_coef_min <- as.vector(coef(lasso_cv, s = "lambda.min"))
lasso_coef_1se <- as.vector(coef(lasso_cv, s = "lambda.1se"))
term_names <- rownames(coef(lasso_cv, s = "lambda.min"))

# Bind into a tidy table and filter for active features
lasso_coef_table <- tibble::tibble(
  term = term_names,
  lambda_min = lasso_coef_min,
  lambda_1se = lasso_coef_1se
) %>%
  # Keep only features that survived soft-thresholding in at least one model
  dplyr::filter(lambda_min != 0 | lambda_1se != 0)

# Render the table for the RMarkdown report
knitr::kable(
  lasso_coef_table,
  digits = 4,
  caption = "Lasso Active Predictors and Coefficients"
)
```

*Application-Specific Interpretation*
The strict $L_1$ penalty at lambda.1se yields a sparse, clinically logical feature set by isolating only age, height, abdom, ankle, and wrist. By applying coordinatewise soft-thresholding, the lasso objective effectively breaks the multicollinearity seen in the OLS baseline, completely zeroing out aggregate mass measurements like weight and hip. Biologically, this indicates that the model relies on abdom as the primary direct proxy for central adiposity, while retaining height, ankle, and wrist to efficiently control for the variance in the patient's underlying skeletal frame size. This proves that once structural bone metrics and abdominal volume are established, an aggregate measure like overall weight becomes mathematically redundant for predicting body fat percentage.

## 2D. Training-based comparison and locked core model

```{r core-comparison, eval=FALSE}
# 2D. Controlled comparison and locked core model

# Helper to extract slopes (excluding intercept) and compute L2 norm
get_l2_norm <- function(fit, s = NULL) {
  if (inherits(fit, "cv.glmnet")) {
    b <- as.matrix(stats::coef(fit, s = s))
  } else {
    b <- as.matrix(stats::coef(fit))
  }
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sqrt(sum(slopes^2))
}

# Compile the comparison table
comparison_table <- tibble::tibble(
  Model = c("OLS", "Ridge", "Ridge", "Lasso", "Lasso"),
  `Tuning Rule` = c("Unpenalized", "lambda.min", "lambda.1se", "lambda.min", "lambda.1se"),
  `CV MSE` = c(
    ols_cv_mse, 
    ridge_err_min, 
    ridge_err_1se, 
    lasso_err_min, 
    lasso_err_1se
  ),
  `L2 Norm` = c(
    get_l2_norm(ols_fit),
    get_l2_norm(ridge_cv, "lambda.min"),
    get_l2_norm(ridge_cv, "lambda.1se"),
    get_l2_norm(lasso_cv, "lambda.min"),
    get_l2_norm(lasso_cv, "lambda.1se")
  ),
  `Nonzero Slopes` = c(
    length(stats::coef(ols_fit)) - 1L,
    count_nonzero(ridge_cv, "lambda.min"),
    count_nonzero(ridge_cv, "lambda.1se"),
    count_nonzero(lasso_cv, "lambda.min"),
    count_nonzero(lasso_cv, "lambda.1se")
  )
)

# Render the table for the report
knitr::kable(
  comparison_table,
  digits = 4,
  caption = "Training-Based Controlled Comparison of Regularization Strategies"
)
```

*1. Decision Questions: lambda.min vs. lambda.1se*
The lambda.min threshold answers an empirical prediction question: which penalty strictly minimizes the cross-validation error? It greedily optimizes for accuracy. Conversely, lambda.1se answers a question of structural stability: what is the heaviest penalty we can apply to simplify the model without suffering a statistically significant loss in predictive power?  

*2. Predictive Performance vs. Interpretability*
Predictive optimization focuses strictly on minimizing the loss surface. As shown in table, Lasso at lambda.min achieves the best predictive performance (CV MSE: 17.3599) but retains 10 slopes, meaning it likely still includes redundant physiological measurements. Interpretability, however, requires exact variable selection. The $L_1$ soft-thresholding mechanism explicitly isolates only the most critical features by forcing weaker, collinear coefficients to exactly zero. 

*3. Locked Core Model Nomination *
Nominated Model:Lasso at lambda.1se.
Justification:
We lock this model based entirely on uncontaminated training data. The dense OLS baseline (14 slopes) exhibits an inflated $L_2$ norm (10.7385), indicating severe multicollinearity. While Ridge regularizes this norm effectively, it cannot perform variable selection. Lasso at lambda.1se provides the optimal clinical solution: it trades a minor increase in CV MSE (to 19.0647) to aggressively reduce the model to exactly 5 non-redundant slopes. This yields a highly interpretable, mathematically stable model that isolates the fundamental biological drivers without overfitting

## 2E. Perturbation or stability check

**Prediction before running the check:**
Prediction before running the check:
  We will duplicate the most mathematically significant structural predictor (e.g., abdom) and add a small amount of random Gaussian noise.
  Ridge Prediction: Because the $L_2$ penalty mathematically minimizes the sum of squared coefficients, it strictly penalizes concentrated weights ($a^2 + a^2 < (2a)^2 + 0^2$). Therefore, ridge regression will split the original coefficient's weight roughly equally between the original predictor and the noisy duplicate to minimize the $L_2$ norm.
  Lasso Prediction: The non-differentiable $L_1$ penalty induces exact sparsity via soft-thresholding. Lasso cannot split weights evenly without increasing the $L_1$ norm constraint. Therefore, lasso will arbitrarily select one of the two highly correlated features to retain and force the coefficient of the other entirely to zero
```{r perturbation, eval=FALSE}
# Ensure strict reproducibility for random noise generation
set.seed(seeds$features)

# Isolate the core biological predictor identified in 2C
target_col <- "abdom" 
if (!target_col %in% colnames(x_train)) target_col <- colnames(x_train)[1]

# Duplicate the predictor and inject small random Gaussian noise
noise <- rnorm(nrow(x_train), mean = 0, sd = 0.01)
x_train_perturbed <- cbind(x_train, perturbed_var = x_train[, target_col] + noise)

# Fit Ridge and Lasso on the strictly locked perturbed training data
ridge_pert <- fit_cv_glmnet(x = x_train_perturbed, y = y_train, alpha = 0, foldid = foldid)
lasso_pert <- fit_cv_glmnet(x = x_train_perturbed, y = y_train, alpha = 1, foldid = foldid)

# Extract coefficients at the strictly regularized lambda.1se threshold
ridge_coefs_pert <- as.vector(stats::coef(ridge_pert, s = "lambda.1se"))
lasso_coefs_pert <- as.vector(stats::coef(lasso_pert, s = "lambda.1se"))
terms_pert <- rownames(stats::coef(ridge_pert))

# Isolate the original and duplicated variables for the report
pert_comparison <- tibble::tibble(
  Term = terms_pert,
  `Ridge (alpha=0)` = ridge_coefs_pert,
  `Lasso (alpha=1)` = lasso_coefs_pert
) %>%
  dplyr::filter(Term %in% c(target_col, "perturbed_var"))

# Render comparison
knitr::kable(
  pert_comparison,
  digits = 4,
  caption = "Coefficient Behavior Under Extreme Collinearity (Perturbation Check)"
)
```

**Observed result and explanation:**
The empirical results perfectly match our theoretical predictions regarding optimization under extreme collinearity.   Ridge ($L_2$) Match: As predicted, the $L_2$ penalty stabilized the model by distributing the coefficient weight almost equally between the original abdom feature (1.9222) and its noisy duplicate (1.9117). This confirms the grouping effect, where the objective minimizes the $L_2$ norm by splitting weights rather than concentrating them.
  Lasso ($L_1$) Match: As predicted, the non-differentiable $L_1$ penalty induced sparsity. The coordinatewise soft-thresholding operator broke the collinearity entirely, assigning the vast majority of the weight to abdom (6.0096) and heavily shrinking the redundant duplicate (0.1054).

# Problem 3. Mathematical mechanisms

## 3A. Ridge and conditioning

*Write the ridge derivation. Define dimensions and explain uniqueness.*

```{r ridge-conditioning, eval=FALSE}
# Use x_train and the fitted ridge lambda.min.
# safe_condition_numbers(x_train, lambda = ...)
```

## 3B. Lasso optimality and soft thresholding

*Derive the zero/nonzero optimality conditions and the orthonormal-design formula.*

## 3C. Connect algebra to evidence

*Connect one specific numerical result to Problem 3A or 3B.*

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research direction:** [weight decay / sparse weights / pruning / dropout]

| Claim | Exact primary source and locator | What it establishes | What it does not establish |
|---|---|---|---|
| [Claim] | [Citation, section/page/equation] | [Support] | [Boundary] |

## 4B. Elastic net on original predictors

```{r original-elastic-net, eval=FALSE}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Fit one cv.glmnet object per alpha with the same foldid.
# Select alpha and lambda using training CV only.
```

## 4C. Fixed ReLU features

```{r random-relu-features, eval=FALSE}
relu <- function(z) pmax(z, 0)
M <- 200L

# TODO:
# 1. Generate A and bias once from seeds$features using the stated distributions.
# 2. Apply the same A and bias to x_train and x_test.
# 3. Fit a scaler to H_train only and apply it to H_test.
# 4. Fit ridge, lasso, and elastic net on H_train with the shared foldid.
```

*State which weights are fixed, which are learned, and how active features are counted.*

## 4D. Locked declaration and final holdout evaluation

> **Locked before viewing `y_test`:** [preprocessing, candidate specifications, selected tuning values, feature seed, nominated deployment model, metrics]

```{r final-holdout-evaluation, eval=FALSE}
# This must be the first analysis chunk that uses y_test for evaluation.
# Generate all fixed candidate predictions and one RMSE/MAE table here.
```

*State whether the holdout evidence supports the predeclared choice. Do not retune.*

## 4E. Deep-learning connection and limitations

*Distinguish penalty, weight decay, output-weight sparsity, pruning, dropout, and a trained deep network. Cite the sources from 4A and discuss at least two limitations.*

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

```text
1. Open a clean R session.
2. [Exact commands and folder layout.]
3. Render GroupXX_ProjectQuiz1.Rmd.
```

## 5B. Recommendation and limitations

*Give the final recommendation, distinguish prediction from interpretation, and state limitations.*

## 5C. AI verification record

| AI-assisted item | How it was checked | Evidence | Modification or rejection |
|---|---|---|---|
| [Item] | [Method] | [Test/source/output] | [Decision] |

# Preflight and session information {-}

```{r preflight, eval=FALSE}
# Add every object required to reproduce the final table and figures.
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", "foldid"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects)) {
  stop(
    "Missing required objects: ",
    paste(missing_objects, collapse = ", ")
  )
}
stopifnot(length(intersect(train_id, test_id)) == 0L)
```

```{r session-info}
sessionInfo()
```
