---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author: "Group 05: [Bùi Trọng Thành - 24125043, Phan Anh Khoa - 24125059, Văn Tuấn Khải - 24125032, Huỳnh Hữu Phúc Minh - 24125068]"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
params:
  group_number: 05
fontsize: 11pt
geometry: margin=1in
---


# Authorship and contribution statement {-}

We confirm that every group member can explain the submitted analysis. AI use is reported in `GroupXX_AI_Log.csv`, and the group checked every AI-assisted item used in this report.

| Student | ID | Main contributions | Verification performed |
|---|---|---|---|
| [Name] | [ID] | [Work] | [Check] |

# Abstract {-}

*Maximum 150 words: prediction problem, methods, main evidence, and recommendation.*

# Reproducible setup {-}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)

group_number <- as.integer(params$group_number)
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```

## Shared helper functions {-}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration, eval=FALSE}
config <- if (group_number %% 2L == 1L) {
  list(
    file = "fat.csv",
    response = "brozek",
    excluded = c("brozek", "siri", "density", "free")
  )
} else {
  list(
    file = "prostate.csv",
    response = "lpsa",
    excluded = "lpsa"
  )
}

data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```

*Define the prediction task and justify every exclusion.*
Target (Response): The prediction target is brozek.  

Candidate Predictors: The available predictors are the remaining non-excluded anatomical measurements (e.g., age, weight, height, adipos, neck, chest, abdom, hip, thigh, knee, ankle, biceps, forearm, and wrist).

Unit of Observation: A single male subject. 

Study Population: The population of men evaluated in the body-fat study described by Johnson (1996).  

Prediction Objective: To predict an individual's body fat percentage (brozek) relying solely on simple, non-invasive anatomical measurements.

Dataset Source: The original study described by Johnson (1996).  

Required Exclusions & Leakage Justification: We are required to explicitly exclude siri, density, and free. These specific variables encode alternative body-fat calculations or are physical quantities derived using body-fat information.

Including any outcome-derived variable fundamentally invalidates the prediction task by introducing target leakage. If included, the regularized models would not learn a generalized statistical relationship from anatomical features; instead, they would merely reverse-engineer a deterministic algebraic formula, rendering any comparison of OLS, ridge, and lasso meaningless.

## 1B. Split and train-only preprocessing

```{r split-and-scale, eval=FALSE}

if (anyNA(analysis_data)) {
  stop("Missing values detected.")
}

split <- split_rows(nrow(analysis_data), train_prop = 0.80, seed = seeds$split)
train_id <- split$train
test_id  <- split$test

# Isolate raw matrices
x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])

# Apply training scalars to both partitions
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test  <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)

# Isolate responses
y_train <- y_raw[train_id]
y_test  <- y_raw[test_id] # REMAIN LOCKED UNTIL PROBLEM 4D

# Shared 5-fold cross-validation allocation
foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

*Report seeds, dimensions, scaling, folds, and leakage safeguards.*
Data Split: We applied an $80/20$ train/test split utilizing the group-specific seed $240205$. This yields a training sample of $n = 201$ observations and a locked holdout sample of $n = 51$ observations.  

Missing-Value Plan: The dataset contains zero missing values. Had missings been detected, any imputation parameters (such as the median or mode) would be calculated exclusively on the $201$ training observations to prevent holdout distribution data from influencing model inputs.  

Scaling Rule: Regularization imposes penalties ($L_1, L_2$) on coefficient magnitudes. To ensure all candidate predictors face equitable shrinkage, we standardize the design matrix. The transformation is computed as  $z_{i,j} = \frac{x_{i,j} - \mu_{train,j}}{\sigma_{train,j}}$. Crucially, $\mu$ and $\sigma$ are derived exclusively from the training matrix. Applying training constants to the test matrix prevents structural target leakage, ensuring no variance or distributional shifts from the holdout set contaminate the model's feature space.  

Cross-Validation Assignment: We defined a single, reproducible 5-fold assignment (foldid) on the training observations using the seed $240305$. Reusing this exact vector across OLS, Ridge, and Lasso guarantees that models are compared on identical out-of-fold samples, removing variance due to random resampling from the model comparison.  

Leakage Prevention Summary: By sequestering y_test, calculating scaling factors solely on $X_{train}$, and locking a shared cross-validation grid, the model acts entirely blind to the holdout data. This isolates the true generalization error for our final evaluation. 

## 1C. Training-data exploration

```{r training-eda, echo=TRUE, message=FALSE, warning=FALSE}
# Isolate training data to strictly prevent leakage
train_data <- analysis_data[train_id, ]

# 1. Response Distribution
p_dist <- ggplot(train_data, aes(x = brozek)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(title = "Distribution of Brozek Body Fat % (Training Set)",
       x = "Brozek (%)", y = "Count")
print(p_dist)

# 2. Focused Relationship / Correlation Table
# Identify top features correlated with the target
cor_matrix <- cor(train_data)
cor_with_target <- sort(abs(cor_matrix[, "brozek"]), decreasing = TRUE)
knitr::kable(head(cor_with_target, 6), 
             col.names = c("Absolute Correlation with Brozek"), digits = 3,
             caption = "Top Correlated Predictors (Training Set)")

# 3. Anomaly Detection
# Scatter plot to identify severe outliers
p_anom <- ggplot(train_data, aes(x = height, y = weight)) +
  geom_point(color = "darkred", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Anomaly Check: Height vs. Weight",
       x = "Height (inches)", y = "Weight (lbs)")
print(p_anom)

# Isolate the mathematically anomalous observation
anomaly_record <- train_data |> filter(height < 40 | weight > 300)
knitr::kable(anomaly_record[, c("brozek", "age", "weight", "height")], 
             caption = "Potential Influential / Anomalous Observations")

# 4. Multicollinearity Detection
# Find highly collinear predictor pairs (r > 0.85) among independent variables
pred_cor <- cor(train_data |> select(-brozek))
pred_cor[lower.tri(pred_cor, diag = TRUE)] <- NA
high_cor <- as.data.frame(as.table(pred_cor)) |>
  filter(abs(Freq) > 0.85) |>
  arrange(desc(abs(Freq)))

knitr::kable(head(high_cor, 5), 
             col.names = c("Predictor 1", "Predictor 2", "Correlation"), digits = 3,
             caption = "Highly Collinear Predictor Pairs (r > 0.85)")
```

*Interpret multicollinearity, one anomaly, and the question regularization will address.*

Response distribution: The brozek target variable is approximately normally distributed and centered near 19%, indicating that ordinary least squares (OLS) and its regularized variants will not require immediate non-linear transformations to achieve stable residuals.

Focused relationship: Torso measurements demonstrate the strongest linear relationship with the target within the training fold, specifically abdom ($r = 0.817$) and adipos ($r = 0.744$).

Anomalous observation: A mathematically severe anomaly exists (height = 29.5 inches, weight = 205.0 lbs), which is almost certainly a data-entry error. Rather than dropping it, this high-leverage point is retained to empirically evaluate how $L_1$ and $L_2$ penalties stabilize estimation when the design matrix contains structural errors compared to the unpenalized OLS baseline.  

Multicollinearity: Severe pairwise correlations are present among spatial measurements, notably weight and hip ($r = 0.940$), and adipos and abdom ($r = 0.932$). In an OLS model, this collinearity inflates the variance of coefficient estimates and drives up the condition number $\kappa_2(G)$ of the Gram matrix $G = \frac{1}{n}X^{\top}X$.

Concluding Question:
Given the severe multicollinearity among body circumferences, can an $L_1$ penalty (lasso) isolate a sparse, interpretable subset of features without sacrificing the predictive stability provided by an $L_2$ penalty (ridge) that smoothly shrinks highly correlated predictors?
# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

```{r ols, eval=FALSE}
# Fit OLS on x_train and y_train. Do not access y_test here.
```

## 2B. Ridge

```{r ridge, eval=FALSE}
# Use fit_cv_glmnet(..., alpha = 0, foldid = foldid).
# Report lambda.min, lambda.1se, paths, coefficients, and CV evidence.
```

## 2C. Lasso

```{r lasso, eval=FALSE}
# Use fit_cv_glmnet(..., alpha = 1, foldid = foldid).
# Report lambda.min, lambda.1se, paths, and nonzero predictors.
```

## 2D. Training-based comparison and locked core model

```{r core-comparison, eval=FALSE}
# Build a comparison from training/CV quantities only.
```

**Core model nominated before holdout evaluation:** [method and reason]

## 2E. Perturbation or stability check

**Prediction before running the check:** [prediction]

```{r perturbation, eval=FALSE}
# Implement one allowed training-only stability check.
```

**Observed result and explanation:** [result]

# Problem 3. Mathematical mechanisms
## 3A. Ridge and conditioning

Starting from the objective function for ridge regression:
$$Q_{\lambda}(b)=\frac{1}{2n}||y-Xb||_{2}^{2}+\frac{\lambda}{2}||b||_{2}^{2}$$

We take the derivative of $Q_{\lambda}(b)$ with respect to the coefficient vector $b$:
$$\nabla_{b}Q_{\lambda}(b)=-\frac{1}{n}X^{\top}(y-Xb)+\lambda b$$

To find the minimizer, we set the gradient to zero:
$$-\frac{1}{n}X^{\top}y+\frac{1}{n}X^{\top}Xb+\lambda b=0$$

Given the definitions of the Gram matrix $G=\frac{1}{n}X^{\top}X$ and the vector $z=\frac{1}{n}X^{\top}y$, we substitute them into the equation:
$$-z+Gb+\lambda b=0$$
$$(G+\lambda I_{p})b=z$$

**Uniqueness explanation:** The Gram matrix $G$ is positive semi-definite. For any penalty parameter $\lambda>0$, the scaled identity matrix $\lambda I_{p}$ is strictly positive definite. Therefore, their sum $(G+\lambda I_{p})$ is always strictly positive definite and invertible. This guarantees that the normal equations have a unique minimizer, given by:
$$\hat{b}_{ridge}=(G+\lambda I_{p})^{-1}z$$

```{r ridge-conditioning}
# Note: ridge_lambda_min must be retrieved from Problem 2B results.
ridge_lambda_min <- 0.6401 

cond_metrics <- safe_condition_numbers(x_train, lambda = ridge_lambda_min)
print(cond_metrics)

## ```{r ridge-conditioning, eval=FALSE}
# Use x_train and the fitted ridge lambda.min.
# safe_condition_numbers(x_train, lambda = ...)
```
## 3B. Lasso optimality and soft thresholding

The objective function for lasso regression is[cite: 1]:
$$R_{\lambda}(b)=\frac{1}{2n}||y-Xb||_{2}^{2}+\lambda||b||_{1}$$

Unlike ridge, the $L_1$ penalty is not differentiable at $b_j = 0$. Therefore, we use subgradients. The optimality condition requires that zero is in the subdifferential of the objective with respect to each coefficient $b_j$:
$$\frac{\partial}{\partial b_j} \left( \frac{1}{2n}||y-Xb||_{2}^{2} \right) + \lambda \partial |b_j| \ni 0$$

Using $G=\frac{1}{n}X^{\top}X$ and $z=\frac{1}{n}X^{\top}y$[cite: 1], the partial derivative of the loss term is $(Gb)_j - z_j$. The subdifferential of $|b_j|$ is $sign(b_j)$ if $b_j \neq 0$, and the interval $[-1, 1]$ if $b_j = 0$.

**1. Optimality conditions for zero and nonzero coefficients:**
*   **Nonzero coefficient ($\hat{b}_j \neq 0$):** The condition is an exact equality:
    $$(G\hat{b})_j - z_j + \lambda \cdot sign(\hat{b}_j) = 0$$
*   **Zero coefficient ($\hat{b}_j = 0$):** The condition becomes an inequality:
    $$|(G\hat{b})_j - z_j| \le \lambda$$
    which means $|z_j - \sum_{k \neq j} G_{jk}\hat{b}_k| \le \lambda$.

**2. Soft-thresholding rule when $G = I_p$:**
When the design matrix is orthonormal, $G = I_p$. Thus, $(G\hat{b})_j = \hat{b}_j$. The optimality conditions simplify to:
$$0 \in \hat{b}_j - z_j + \lambda \partial |\hat{b}_j|$$
*   If $\hat{b}_j > 0$: $\hat{b}_j - z_j + \lambda = 0 \implies \hat{b}_j = z_j - \lambda$ (requires $z_j > \lambda$).
*   If $\hat{b}_j < 0$: $\hat{b}_j - z_j - \lambda = 0 \implies \hat{b}_j = z_j + \lambda$ (requires $z_j < -\lambda$).
*   If $\hat{b}_j = 0$: $|z_j| \le \lambda$.

Combining these three cases yields the soft-thresholding rule[cite: 1]:
$$\hat{b}_{j}=sign(z_{j})(|z_{j}|-\lambda)_{+}$$

**3. Why Lasso produces exact zeros compared to Ridge:**
The lasso optimality condition at $\hat{b}_j = 0$ is a region: the gradient of the unpenalized loss must simply fall within the interval $[-\lambda, \lambda]$. Because this is a wide interval of width $2\lambda$, there is a high probability that the condition is met exactly, pinning the coefficient to absolute zero[cite: 1]. 
In contrast, ridge regression is globally differentiable with a penalty gradient of $\lambda \hat{b}_j$. The ridge optimality condition is $(G + \lambda I_p)\hat{b} = z$. Unless $z$ happens to be exactly zero, the ridge coefficients are only shrunk asymptotically towards zero but rarely hit exactly zero.
## 3B. Lasso optimality and soft thresholding Task RequireMent

*Derive the zero/nonzero optimality conditions and the orthonormal-design formula.*

## 3C. Connect algebra to evidence

*Connect one specific numerical result to Problem 3A or 3B.*

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research direction:** [weight decay / sparse weights / pruning / dropout]

| Claim | Exact primary source and locator | What it establishes | What it does not establish |
|---|---|---|---|
| [Claim] | [Citation, section/page/equation] | [Support] | [Boundary] |

## 4B. Elastic net on original predictors

```{r original-elastic-net, eval=FALSE}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Fit one cv.glmnet object per alpha with the same foldid.
# Select alpha and lambda using training CV only.
```

## 4C. Fixed ReLU features

```{r random-relu-features, eval=FALSE}
relu <- function(z) pmax(z, 0)
M <- 200L

# TODO:
# 1. Generate A and bias once from seeds$features using the stated distributions.
# 2. Apply the same A and bias to x_train and x_test.
# 3. Fit a scaler to H_train only and apply it to H_test.
# 4. Fit ridge, lasso, and elastic net on H_train with the shared foldid.
```

*State which weights are fixed, which are learned, and how active features are counted.*

## 4D. Locked declaration and final holdout evaluation

> **Locked before viewing `y_test`:** [preprocessing, candidate specifications, selected tuning values, feature seed, nominated deployment model, metrics]

```{r final-holdout-evaluation, eval=FALSE}
# This must be the first analysis chunk that uses y_test for evaluation.
# Generate all fixed candidate predictions and one RMSE/MAE table here.
```

*State whether the holdout evidence supports the predeclared choice. Do not retune.*

## 4E. Deep-learning connection and limitations

*Distinguish penalty, weight decay, output-weight sparsity, pruning, dropout, and a trained deep network. Cite the sources from 4A and discuss at least two limitations.*

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

```text
1. Open a clean R session.
2. [Exact commands and folder layout.]
3. Render GroupXX_ProjectQuiz1.Rmd.
```

## 5B. Recommendation and limitations

*Give the final recommendation, distinguish prediction from interpretation, and state limitations.*

## 5C. AI verification record

| AI-assisted item | How it was checked | Evidence | Modification or rejection |
|---|---|---|---|
| [Item] | [Method] | [Test/source/output] | [Decision] |

# Preflight and session information {-}

```{r preflight, eval=FALSE}
# Add every object required to reproduce the final table and figures.
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", "foldid"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects)) {
  stop(
    "Missing required objects: ",
    paste(missing_objects, collapse = ", ")
  )
}
stopifnot(length(intersect(train_id, test_id)) == 0L)
```

```{r session-info}
sessionInfo()
```
