---
title: "ASESII 24A02 Project-Based Quiz 1"
author: |
  | Group 05:
  | Bùi Trọng Thành (24125043)
  | Phan Anh Khoa (24125059)
  | Văn Tuấn Khải (24125032)
  | Huỳnh Hữu Phúc Minh (24125068)
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
  word_document:
    toc: true
    toc_depth: '3'
subtitle: Ridge, Lasso, and Regularization for Neural Features
bibliography: references.bib
link-citations: true
params:
  group_number: 5
fontsize: 11pt
geometry: margin=1in
---


# Authorship and contribution statement {-}

We confirm that every group member can explain the submitted analysis. AI use is reported in section 5C of this report, and the group checked every AI-assisted item used in this report.

| Student | Student ID | Main Contributions | Verification Performed |
|---------|-----------|---------------------|-------------------------|
| Thanh | [24125043] | Prediction design, leakage control safeguards (P1), and OLS/Ridge baseline training (P2A–2B). | Audited the fixed ReLU feature generation and Elastic Net cross-validation (P4) to confirm fixed vs. learned parameters. |
| Khoa | [24125059] | Lasso optimization, perturbation checks (P2C–2E), and theoretical mechanism derivations (P3). | Validated the 80/20 splitting logic and ensured strict separation of test responses (P1). |
| Minh | [24125068] | Elastic net tuning and implementation of the fixed non-linear ReLU feature representation (P4A–4C). | Verified the exact coordinatewise optimality conditions and eigenvalue conditioning κ₂(G+λI_p) (P3). |
| Khai | [24125032] | Deep learning literature synthesis (P4E), locked holdout evaluation (P4D), and reproducible setup (P5). | Reviewed all cross-validation folds (foldid) for consistency across OLS, Ridge, and Lasso (P2). |

# Abstract {-}

This study evaluates regularized regression (Ridge, Lasso, Elastic Net) to predict body fat percentage (brozek) from anatomical measurements. We compare classical $L_1$ and $L_2$ penalties against a fixed 200-dimensional ReLU feature space, establishing a controlled bridge to neural architectures. Uncompromising data hygiene was maintained via a locked $80/20$ split and training-exclusive scaling. While unpenalized OLS yielded a dangerously ill-conditioned Gram matrix, regularization stabilized the eigenvalue conditioning $\kappa_2(G+\lambda I_p)$. Despite the Elastic Net ReLU model achieving the lowest holdout RMSE (4.1310), we recommend deploying Lasso at lambda.1se (RMSE 4.4070). Its coordinatewise soft-thresholding mechanism aggressively broke multicollinearity, isolating exactly five fundamental clinical predictors. This yields an optimally stable, highly interpretable model. Limitations include the ReLU representation's high variance to stochastic initialization and its fundamental lack of gradient flow for hierarchical learning

# Reproducible setup {-}

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(width = 60)

# In setup chunk:
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$linewidth)) {
    x <- knitr:::split_lines(x)
    if (any(nchar(x) > n)) x <- strwrap(x, width = n)
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
required_packages <- c("tidyverse", "glmnet", "broom", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)
library(tibble)

group_number <- as.integer(params$group_number)
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```

## Shared helper functions {-}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors
```{r data-configuration}
config <- if (group_number %% 2L == 1L) {
  list(
    file = "fat.csv",
    response = "brozek",
    excluded = c("brozek", "siri", "density", "free")
  )
} else {
  list(
    file = "prostate.csv",
    response = "lpsa",
    excluded = "lpsa"
  )
}

data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```

*Define the prediction task and justify every exclusion.*
Target (Response): The prediction target is brozek.  

Candidate Predictors: The available predictors are the remaining non-excluded anatomical measurements (e.g., age, weight, height, adipos, neck, chest, abdom, hip, thigh, knee, ankle, biceps, forearm, and wrist).

Unit of Observation: A single male subject. 

Study Population: The population of men evaluated in the body-fat study described by Johnson (1996).  

Prediction Objective: To predict an individual's body fat percentage (brozek) relying solely on simple, non-invasive anatomical measurements.

Dataset Source: The original study described by Johnson (1996).  

Required Exclusions & Leakage Justification: We are required to explicitly exclude siri, density, and free. These specific variables encode alternative body-fat calculations or are physical quantities derived using body-fat information.

Including any outcome-derived variable fundamentally invalidates the prediction task by introducing target leakage. If included, the regularized models would not learn a generalized statistical relationship from anatomical features; instead, they would merely reverse-engineer a deterministic algebraic formula, rendering any comparison of OLS, ridge, and lasso meaningless.

## 1B. Split and train-only preprocessing
```{r split-and-scale}

if (anyNA(analysis_data)) {
  stop("Missing values detected.")
}

split <- split_rows(nrow(analysis_data), train_prop = 0.80, seed = seeds$split)
train_id <- split$train
test_id  <- split$test

# Isolate raw matrices
x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])

# Apply training scalars to both partitions
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test  <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)

# Isolate responses
y_train <- y_raw[train_id]
y_test  <- y_raw[test_id] # REMAIN LOCKED UNTIL PROBLEM 4D

# Shared 5-fold cross-validation allocation
foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

*Report seeds, dimensions, scaling, folds, and leakage safeguards.*
Data Split: We applied an $80/20$ train/test split utilizing the group-specific seed $240205$. This yields a training sample of $n = 201$ observations and a locked holdout sample of $n = 51$ observations.  

Missing-Value Plan: The dataset contains zero missing values. Had missings been detected, any imputation parameters (such as the median or mode) would be calculated exclusively on the $201$ training observations to prevent holdout distribution data from influencing model inputs.  

Scaling Rule: Regularization imposes penalties ($L_1, L_2$) on coefficient magnitudes. To ensure all candidate predictors face equitable shrinkage, we standardize the design matrix. The transformation is computed as  $z_{i,j} = \frac{x_{i,j} - \mu_{train,j}}{\sigma_{train,j}}$. Crucially, $\mu$ and $\sigma$ are derived exclusively from the training matrix. Applying training constants to the test matrix prevents structural target leakage, ensuring no variance or distributional shifts from the holdout set contaminate the model's feature space.  

Cross-Validation Assignment: We defined a single, reproducible 5-fold assignment (foldid) on the training observations using the seed $240305$. Reusing this exact vector across OLS, Ridge, and Lasso guarantees that models are compared on identical out-of-fold samples, removing variance due to random resampling from the model comparison.  

Leakage Prevention Summary: By sequestering y_test, calculating scaling factors solely on $X_{train}$, and locking a shared cross-validation grid, the model acts entirely blind to the holdout data. This isolates the true generalization error for our final evaluation. 

## 1C. Training-data exploration

```{r training-eda, echo=TRUE, message=FALSE, warning=FALSE}
# Isolate training data to strictly prevent leakage
train_data <- analysis_data[train_id, ]

# 1. Response Distribution
p_dist <- ggplot(train_data, aes(x = brozek)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(title = "Distribution of Brozek Body Fat % (Training Set)",
       x = "Brozek (%)", y = "Count")
print(p_dist)

# 2. Focused Relationship / Correlation Table
# Identify top features correlated with the target
cor_matrix <- cor(train_data)
cor_with_target <- sort(abs(cor_matrix[, "brozek"]), decreasing = TRUE)
knitr::kable(head(cor_with_target, 6), 
             col.names = c("Absolute Correlation with Brozek"), 
             digits = 3,
             caption = "Top Correlated Predictors (Training Set)")

# 3. Anomaly Detection
# Scatter plot to identify severe outliers
p_anom <- ggplot(train_data, aes(x = height, y = weight)) +
  geom_point(color = "darkred", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Anomaly Check: Height vs. Weight",
       x = "Height (inches)", y = "Weight (lbs)")
print(p_anom)

# Isolate the mathematically anomalous observation
anomaly_record <- train_data |> filter(height < 40 | weight > 300)
knitr::kable(anomaly_record[, c("brozek", "age", "weight", "height")], 
             caption = "Potential Influential / Anomalous Observations")

# 4. Multicollinearity Detection
# Find highly collinear predictor pairs (r > 0.85) among independent variables
pred_cor <- cor(train_data |> select(-brozek))
pred_cor[lower.tri(pred_cor, diag = TRUE)] <- NA
high_cor <- as.data.frame(as.table(pred_cor)) |>
  filter(abs(Freq) > 0.85) |>
  arrange(desc(abs(Freq)))

knitr::kable(head(high_cor, 5), 
             col.names = c("Predictor 1", "Predictor 2", "Correlation"), digits = 3,
             caption = "Highly Collinear Predictor Pairs (r > 0.85)")
```

*Interpret multicollinearity, one anomaly, and the question regularization will address.*

Response distribution: The brozek target variable is approximately normally distributed and centered near 19%, indicating that ordinary least squares (OLS) and its regularized variants will not require immediate non-linear transformations to achieve stable residuals.

Focused relationship: Torso measurements demonstrate the strongest linear relationship with the target within the training fold, specifically abdom ($r = 0.817$) and adipos ($r = 0.744$).

Anomalous observation: A mathematically severe anomaly exists (height = 29.5 inches, weight = 205.0 lbs), which is almost certainly a data-entry error. Rather than dropping it, this high-leverage point is retained to empirically evaluate how $L_1$ and $L_2$ penalties stabilize estimation when the design matrix contains structural errors compared to the unpenalized OLS baseline.  

Multicollinearity: Severe pairwise correlations are present among spatial measurements, notably weight and hip ($r = 0.940$), and adipos and abdom ($r = 0.932$). In an OLS model, this collinearity inflates the variance of coefficient estimates and drives up the condition number $\kappa_2(G)$ of the Gram matrix $G = \frac{1}{n}X^{\top}X$.

Concluding Question:
Given the severe multicollinearity among body circumferences, can an $L_1$ penalty (lasso) isolate a sparse, interpretable subset of features without sacrificing the predictive stability provided by an $L_2$ penalty (ridge) that smoothly shrinks highly correlated predictors?

# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

```{r ols}
# 1. Fit OLS Model
df_train <- data.frame(y = y_train, x_train)
ols_fit <- lm(y ~ ., data = df_train)

# 2. Extract and Report Coefficients (Idiomatic tidyverse/broom)
ols_coef_table <- broom::tidy(ols_fit)

# Output table for the report
knitr::kable(
  ols_coef_table, 
  digits = 4, 
  caption = "OLS Baseline Coefficients"
)

# 3. Calculate Errors (MSE)
ols_train_mse <- mean(resid(ols_fit)^2)

ols_cv_mse <- purrr::map_dbl(1:5, function(k) {
  val_mask <- (foldid == k)
  
  # Strict isolation: train only on out-of-fold data
  fit_k <- lm(y ~ ., data = df_train[!val_mask, ])
  preds_k <- predict(fit_k, newdata = df_train[val_mask, ])
  
  # Calculate validation MSE for fold k
  mean((df_train$y[val_mask] - preds_k)^2)
}) %>% mean()

# 4. Report Errors
cat(sprintf("Training MSE: %.4f\n", ols_train_mse))
cat(sprintf("5-Fold CV MSE: %.4f\n", ols_cv_mse))

# 5. Diagnostic Plot
plot(ols_fit, which = 1, main = "OLS Residuals vs Fitted")
```

Interpretation of OLS Baseline:
The OLS model exhibits severe instability due to multicollinearity among the physical measurements. Evidence of this ill-conditioning includes inflated standard errors (e.g., weight $SE \approx 1.638$) and physiologically counterintuitive coefficient signs (e.g., a negative coefficient for weight). Because the highly correlated predictors cause the eigenvalues of the Gram matrix to approach zero, the variance of the unpenalized $\hat{\beta}$ estimates explodes. We retain all features as instructed; however, this instability necessitates the $L_1$ and $L_2$ regularization approaches explored in subsequent sections.

## 2B. Ridge

```{r ridge}
# Fit Ridge CV using shared folds to prevent data leakage
ridge_cv <- fit_cv_glmnet(x_train, y_train, alpha = 0, foldid = foldid)

# Extract lambdas
ridge_lambda_min <- ridge_cv$lambda.min
ridge_lambda_1se <- ridge_cv$lambda.1se

# Extract CV errors
ridge_err_min <- ridge_cv$cvm[ridge_cv$lambda == ridge_lambda_min]
ridge_err_1se <- ridge_cv$cvm[ridge_cv$lambda == ridge_lambda_1se]

# Bind coefficients for clean reporting
ridge_coefs <- tibble::tibble(
  term = rownames(coef(ridge_cv)),
  lambda_min = as.vector(coef(ridge_cv, s = "lambda.min")),
  lambda_1se = as.vector(coef(ridge_cv, s = "lambda.1se"))
)

# Output formatting for the report
cat(sprintf("lambda.min: %.4f (CV MSE: %.4f)\n", ridge_lambda_min, ridge_err_min))
cat(sprintf("lambda.1se: %.4f (CV MSE: %.4f)\n", ridge_lambda_1se, ridge_err_1se))

knitr::kable(
  ridge_coefs,
  digits = 4,
  caption = "Ridge Coefficients at lambda.min and lambda.1se"
)

# Visualizations: CV Curve and Coefficient Paths
par(mfrow = c(1, 2))
plot(ridge_cv)
title("Ridge CV Error", line = 2.5)

plot(ridge_cv$glmnet.fit, xvar = "lambda", label = TRUE)
abline(v = log(c(ridge_lambda_min, ridge_lambda_1se)), 
       lty = 2, col = c("blue", "red"))
title("Ridge Coefficient Paths", line = 2.5)
par(mfrow = c(1, 1))
```

Interpretation of Ridge Regression Results:

*Tuning and Validation Error*:
Using 5-fold cross-validation on the locked training data, the ridge regression model identified an optimal penalty of lambda.min (CV MSE: 19.7095). Applying the stricter one-standard-error rule yields lambda.1se (CV MSE: 22.1846), which applies heavier regularization in exchange for a slightly higher training estimation error.

*Behavior of Correlated Predictors ($L_2$ Mechanism)*:
The $L_2$ penalty in the ridge objective function, defined as $\frac{\lambda}{2} \vert{}\vert{}b\vert{}\vert{}_2^2$, heavily penalizes large individual coefficients. Consequently, when presented with highly correlated physiological predictors (e.g., weight, abdom, chest), ridge regression exhibits a "grouping effect." Rather than assigning a massive weight to one predictor and zero to another, the $L_2$ norm mathematically prefers to distribute the coefficients somewhat evenly among correlated features, as $a^2 + a^2 < (2a)^2 + 0^2$.
This stabilization is highly visible in the coefficient shifts from lambda.min to lambda.1se. For instance, the heavily inflated abdom coefficient shrinks from 5.0971 to 2.5770. Furthermore, the sign-flipping observed in the ill-conditioned OLS baseline is corrected; the coefficient for weight transitions from a biologically counterintuitive negative value (-0.2324) at lambda.min to a logical positive value (0.3160) at the more heavily penalized lambda.1se.

*Lack of Exact Sparsity*:
It is critical to note that while the $L_2$ penalty shrinks coefficients toward zero asymptotically, it lacks a coordinatewise soft-thresholding mechanism. Therefore, ridge regression does not perform variable selection. Numerically tiny coefficients, such as knee (-0.0032 at lambda.min), remain mathematically active in the prediction equation and must not be interpreted as exact zeros.

## 2C. Lasso

```{r lasso}
# 1. Fit Lasso CV using shared folds
lasso_cv <- fit_cv_glmnet(x = x_train, y = y_train, alpha = 1, foldid = foldid)

# 2. Extract optimal lambdas and their CV errors
lasso_lambda_min <- lasso_cv$lambda.min
lasso_lambda_1se <- lasso_cv$lambda.1se

lasso_err_min <- lasso_cv$cvm[lasso_cv$lambda == lasso_lambda_min]
lasso_err_1se <- lasso_cv$cvm[lasso_cv$lambda == lasso_lambda_1se]

# 3. Custom extraction of exact nonzero feature names
get_active_features <- function(fit, s) {
  # Extract sparse matrix of coefficients
  b <- as.matrix(coef(fit, s = s))
  # Subset to non-zero, non-intercept terms
  active <- rownames(b)[b[, 1] != 0 & rownames(b) != "(Intercept)"]
  return(active)
}

lasso_active_min <- get_active_features(lasso_cv, "lambda.min")
lasso_active_1se <- get_active_features(lasso_cv, "lambda.1se")

# 4. Formatted output for the RMarkdown report
cat(sprintf("lambda.min: %.4f (CV MSE: %.4f)\n", lasso_lambda_min, lasso_err_min))
cat("Active predictors (lambda.min):\n", 
    paste(strwrap(paste(lasso_active_min, collapse = ", "), width = 70), 
          collapse = "\n"), "\n\n")
cat(sprintf("lambda.1se: %.4f (CV MSE: %.4f)\n", lasso_lambda_1se, lasso_err_1se))
cat("Active predictors (lambda.1se):\n", 
    paste(strwrap(paste(lasso_active_1se, collapse = ", "), width = 70), 
          collapse = "\n"), "\n")

# 5. Visualizations
par(mfrow = c(1, 2))

plot(lasso_cv)
title("Lasso CV Error", line = 2.5)

plot(lasso_cv$glmnet.fit, xvar = "lambda", label = TRUE)
abline(v = log(c(lasso_lambda_min, lasso_lambda_1se)), lty = 2, 
       col = c("blue", "red"))
title("Lasso Coefficient Paths", line = 2.5)

par(mfrow = c(1, 1))

# Extract coefficients as vectors
lasso_coef_min <- as.vector(coef(lasso_cv, s = "lambda.min"))
lasso_coef_1se <- as.vector(coef(lasso_cv, s = "lambda.1se"))
term_names <- rownames(coef(lasso_cv, s = "lambda.min"))

# Bind into a tidy table and filter for active features
lasso_coef_table <- tibble::tibble(
  term = term_names,
  lambda_min = lasso_coef_min,
  lambda_1se = lasso_coef_1se
) %>%
  # Keep only features that survived soft-thresholding in at least one model
  dplyr::filter(lambda_min != 0 | lambda_1se != 0)

# Render the table for the RMarkdown report
knitr::kable(
  lasso_coef_table,
  digits = 4,
  caption = "Lasso Active Predictors and Coefficients"
)
```

*Application-Specific Interpretation*
The strict $L_1$ penalty at lambda.1se yields a sparse, clinically logical feature set by isolating only age, height, abdom, ankle, and wrist. By applying coordinatewise soft-thresholding, the lasso objective effectively breaks the multicollinearity seen in the OLS baseline, completely zeroing out aggregate mass measurements like weight and hip. Biologically, this indicates that the model relies on abdom as the primary direct proxy for central adiposity, while retaining height, ankle, and wrist to efficiently control for the variance in the patient's underlying skeletal frame size. This proves that once structural bone metrics and abdominal volume are established, an aggregate measure like overall weight becomes mathematically redundant for predicting body fat percentage.

## 2D. Training-based comparison and locked core model

```{r core-comparison}
# 2D. Controlled comparison and locked core model

# Helper to extract slopes (excluding intercept) and compute L2 norm
get_l2_norm <- function(fit, s = NULL) {
  if (inherits(fit, "cv.glmnet")) {
    b <- as.matrix(stats::coef(fit, s = s))
  } else {
    b <- as.matrix(stats::coef(fit))
  }
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sqrt(sum(slopes^2))
}

# Compile the comparison table
comparison_table <- tibble::tibble(
  Model = c("OLS", "Ridge", "Ridge", "Lasso", "Lasso"),
  `Tuning Rule` = c("Unpenalized", "lambda.min", "lambda.1se", 
                    "lambda.min", "lambda.1se"),
  `CV MSE` = c(
    ols_cv_mse, 
    ridge_err_min, 
    ridge_err_1se, 
    lasso_err_min, 
    lasso_err_1se
  ),
  `L2 Norm` = c(
    get_l2_norm(ols_fit),
    get_l2_norm(ridge_cv, "lambda.min"),
    get_l2_norm(ridge_cv, "lambda.1se"),
    get_l2_norm(lasso_cv, "lambda.min"),
    get_l2_norm(lasso_cv, "lambda.1se")
  ),
  `Nonzero Slopes` = c(
    length(stats::coef(ols_fit)) - 1L,
    count_nonzero(ridge_cv, "lambda.min"),
    count_nonzero(ridge_cv, "lambda.1se"),
    count_nonzero(lasso_cv, "lambda.min"),
    count_nonzero(lasso_cv, "lambda.1se")
  )
)

# Render the table for the report
knitr::kable(
  comparison_table,
  digits = 4,
  caption = "Training-Based Controlled Comparison of Regularization Strategies"
)
```

*1. Decision Questions: lambda.min vs. lambda.1se*
The lambda.min threshold answers an empirical prediction question: which penalty strictly minimizes the cross-validation error? It greedily optimizes for accuracy. Conversely, lambda.1se answers a question of structural stability: what is the heaviest penalty we can apply to simplify the model without suffering a statistically significant loss in predictive power?  

*2. Predictive Performance vs. Interpretability*
Predictive optimization focuses strictly on minimizing the loss surface. As shown in table, Lasso at lambda.min achieves the best predictive performance (CV MSE: 17.3599) but retains 10 slopes, meaning it likely still includes redundant physiological measurements. Interpretability, however, requires exact variable selection. The $L_1$ soft-thresholding mechanism explicitly isolates only the most critical features by forcing weaker, collinear coefficients to exactly zero. 

*3. Locked Core Model Nomination *
Nominated Model:Lasso at lambda.1se.
Justification:
We lock this model based entirely on uncontaminated training data. The dense OLS baseline (14 slopes) exhibits an inflated $L_2$ norm (10.7385), indicating severe multicollinearity. While Ridge regularizes this norm effectively, it cannot perform variable selection. Lasso at lambda.1se provides the optimal clinical solution: it trades a minor increase in CV MSE (to 19.0647) to aggressively reduce the model to exactly 5 non-redundant slopes. This yields a highly interpretable, mathematically stable model that isolates the fundamental biological drivers without overfitting

## 2E. Perturbation or stability check

**Prediction before running the check:**
Prediction before running the check:
  We will duplicate the most mathematically significant structural predictor (e.g., abdom) and add a small amount of random Gaussian noise.
  Ridge Prediction: Because the $L_2$ penalty mathematically minimizes the sum of squared coefficients, it strictly penalizes concentrated weights ($a^2 + a^2 < (2a)^2 + 0^2$). Therefore, ridge regression will split the original coefficient's weight roughly equally between the original predictor and the noisy duplicate to minimize the $L_2$ norm.
  Lasso Prediction: The non-differentiable $L_1$ penalty induces exact sparsity via soft-thresholding. Lasso cannot split weights evenly without increasing the $L_1$ norm constraint. Therefore, lasso will arbitrarily select one of the two highly correlated features to retain and force the coefficient of the other entirely to zero
```{r perturbation}
# Ensure strict reproducibility for random noise generation
set.seed(seeds$features)

# Isolate the core biological predictor identified in 2C
target_col <- "abdom" 
if (!target_col %in% colnames(x_train)) target_col <- colnames(x_train)[1]

# Duplicate the predictor and inject small random Gaussian noise
noise <- rnorm(nrow(x_train), mean = 0, sd = 0.01)
x_train_perturbed <- cbind(x_train, perturbed_var = x_train[, target_col] + noise)

# Fit Ridge and Lasso on the strictly locked perturbed training data
ridge_pert <- fit_cv_glmnet(x = x_train_perturbed, y = y_train, 
                            alpha = 0, foldid = foldid)
lasso_pert <- fit_cv_glmnet(x = x_train_perturbed, y = y_train, 
                            alpha = 1, foldid = foldid)

# Extract coefficients at the strictly regularized lambda.1se threshold
ridge_coefs_pert <- as.vector(stats::coef(ridge_pert, s = "lambda.1se"))
lasso_coefs_pert <- as.vector(stats::coef(lasso_pert, s = "lambda.1se"))
terms_pert <- rownames(stats::coef(ridge_pert))

# Isolate the original and duplicated variables for the report
pert_comparison <- tibble::tibble(
  Term = terms_pert,
  `Ridge (alpha=0)` = ridge_coefs_pert,
  `Lasso (alpha=1)` = lasso_coefs_pert
) %>%
  dplyr::filter(Term %in% c(target_col, "perturbed_var"))

# Render comparison
knitr::kable(
  pert_comparison,
  digits = 4,
  caption = "Coefficient Behavior Under Extreme Collinearity (Perturbation Check)"
)
```

**Observed result and explanation:**
The empirical results perfectly match our theoretical predictions regarding optimization under extreme collinearity.   Ridge ($L_2$) Match: As predicted, the $L_2$ penalty stabilized the model by distributing the coefficient weight almost equally between the original abdom feature (1.9222) and its noisy duplicate (1.9117). This confirms the grouping effect, where the objective minimizes the $L_2$ norm by splitting weights rather than concentrating them.
  Lasso ($L_1$) Match: As predicted, the non-differentiable $L_1$ penalty induced sparsity. The coordinatewise soft-thresholding operator broke the collinearity entirely, assigning the vast majority of the weight to abdom (6.0096) and heavily shrinking the redundant duplicate (0.1054).

# Problem 3. Mathematical mechanisms
## 3A. Ridge and conditioning

Starting from the objective function for ridge regression:
$$Q_{\lambda}(b)=\frac{1}{2n}||y-Xb||_{2}^{2}+\frac{\lambda}{2}||b||_{2}^{2}$$
We take the derivative of $Q_{\lambda}(b)$ with respect to the coefficient vector $b$:
$$\nabla_{b}Q_{\lambda}(b)=-\frac{1}{n}X^{\top}(y-Xb)+\lambda b$$

To find the minimizer, we set the gradient to zero:
$$-\frac{1}{n}X^{\top}y+\frac{1}{n}X^{\top}Xb+\lambda b=0$$

Given the definitions of the Gram matrix $G=\frac{1}{n}X^{\top}X$ and the vector $z=\frac{1}{n}X^{\top}y$, we substitute them into the equation:
$$-z+Gb+\lambda b=0$$
$$(G+\lambda I_{p})b=z$$

**Uniqueness explanation:** The Gram matrix $G$ is positive semi-definite. For any penalty parameter $\lambda>0$, the scaled identity matrix $\lambda I_{p}$ is strictly positive definite. Therefore, their sum $(G+\lambda I_{p})$ is always strictly positive definite and invertible. This guarantees that the normal equations have a unique minimizer, given by:
$$\hat{b}_{ridge}=(G+\lambda I_{p})^{-1}z$$


```{r ridge-conditioning}
# Retrieve lambda.min from the fitted Ridge model in Problem 2B
ridge_lambda <- ridge_cv$lambda.min 

# Calculate the Gram matrix G = (1/n) * X^T * X
n <- nrow(x_train)
G <- (1/n) * t(x_train) %*% x_train

# Extract eigenvalues
eigenvalues <- eigen(G)$values
mu_max <- max(eigenvalues)
mu_min <- min(eigenvalues)

# 1. Unpenalized Condition Number: k2(G)
# Taking it to be infinite when the smallest eigenvalue is zero 
# up to numerical tolerance
tol <- 1e-10
if (mu_min < tol) {
  kappa_ols <- Inf
} else {
  kappa_ols <- mu_max / mu_min
}

# 2. Penalized Condition Number: k2(G + lambda*I_p)
kappa_ridge <- (mu_max + ridge_lambda) / (mu_min + ridge_lambda)

# Output the results
cat(sprintf("Largest Eigenvalue (mu_max): %f\n", mu_max))
cat(sprintf("Smallest Eigenvalue (mu_min): %f\n", mu_min))
cat(sprintf("Unpenalized Condition Number k2(G): %f\n", kappa_ols))
cat(sprintf("Ridge Condition Number k2(G + lambda*I): %f\n", kappa_ridge))
```
### Explain what changed geometrically and numerically:
*Numerically*: Extreme multicollinearity among spatial measurements (e.g., weight, abdom, chest) causes the unpenalized design matrix to be ill-conditioned. The smallest eigenvalue $\mu_{min}$ approaches zero, causing the OLS condition number $\kappa_2(G) = \frac{\mu_{max}}{\mu_{min}}$ to explode toward infinity. This ill-conditioning makes matrix inversion highly unstable, resulting in the inflated coefficient variance observed in Problem 2A. By adding the $L_2$ penalty, we shift every eigenvalue strictly upward by $\lambda$. The penalized condition number evaluates to $\frac{\mu_{max} + \lambda}{\mu_{min} + \lambda}$. Because $\lambda > 0$, the denominator is strictly bounded away from zero, which drastically reduces $\kappa_2$ and guarantees a computationally stable matrix inversion.

*Geometrically*: The eigenvalues of $G$ represent the curvature of the loss function along its principal axes. An eigenvalue near zero indicates that the unpenalized OLS loss surface is an extremely flat, elongated valley. In this flat valley, infinitely many coefficient combinations yield nearly identical training errors, making the optimization highly sensitive to noise. Adding $\lambda I_p$ expands the curvature along all axes by exactly $\lambda$. This geometrically reshapes the flat valley into a strictly convex, well-rounded bowl, ensuring that the optimization perfectly converges to a single, stable, and unique global minimum.

## 3B. Lasso optimality and soft thresholding

The objective function for lasso regression is:
$$R_{\lambda}(b)=\frac{1}{2n}||y-Xb||_{2}^{2}+\lambda||b||_{1}$$

Unlike ridge, the $L_1$ penalty is not differentiable at $b_j = 0$. Therefore, we use subgradients. The optimality condition requires that zero is in the subdifferential of the objective with respect to each coefficient $b_j$:
$$\frac{\partial}{\partial b_j} \left( \frac{1}{2n}||y-Xb||_{2}^{2} \right) + \lambda \partial |b_j| \ni 0$$

Using $G=\frac{1}{n}X^{\top}X$ and $z=\frac{1}{n}X^{\top}y$, the partial derivative of the loss term is $(Gb)_j - z_j$. The subdifferential of $|b_j|$ is $sign(b_j)$ if $b_j \neq 0$, and the interval $[-1, 1]$ if $b_j = 0$.

**1. Optimality conditions for zero and nonzero coefficients:**

*   **Nonzero coefficient ($\hat{b}_j \neq 0$):** The condition is an exact equality:
    $$(G\hat{b})_j - z_j + \lambda \cdot sign(\hat{b}_j) = 0$$
*   **Zero coefficient ($\hat{b}_j = 0$):** The condition becomes an inequality:
    $$|(G\hat{b})_j - z_j| \le \lambda$$
    which means $|z_j - \sum_{k \neq j} G_{jk}\hat{b}_k| \le \lambda$.

**2. Soft-thresholding rule when $G = I_p$:**
When the design matrix is orthonormal, $G = I_p$. Thus, $(G\hat{b})_j = \hat{b}_j$. The optimality conditions simplify to:
$$0 \in \hat{b}_j - z_j + \lambda \partial |\hat{b}_j|$$
*   If $\hat{b}_j > 0$: $\hat{b}_j - z_j + \lambda = 0 \implies \hat{b}_j = z_j - \lambda$ (requires $z_j > \lambda$).

*   If $\hat{b}_j < 0$: $\hat{b}_j - z_j - \lambda = 0 \implies \hat{b}_j = z_j + \lambda$ (requires $z_j < -\lambda$).

*   If $\hat{b}_j = 0$: $|z_j| \le \lambda$.

Combining these three cases yields the soft-thresholding rule:
$$\hat{b}_{j}=sign(z_{j})(|z_{j}|-\lambda)_{+}$$

**3. Why Lasso produces exact zeros compared to Ridge:**
The lasso optimality condition at $\hat{b}_j = 0$ is a region: the gradient of the unpenalized loss must simply fall within the interval $[-\lambda, \lambda]$. Because this is a wide interval of width $2\lambda$, there is a high probability that the condition is met exactly, pinning the coefficient to absolute zero. 
In contrast, ridge regression is globally differentiable with a penalty gradient of $\lambda \hat{b}_j$. The ridge optimality condition is $(G + \lambda I_p)\hat{b} = z$. Unless $z$ happens to be exactly zero, the ridge coefficients are only shrunk asymptotically towards zero but rarely hit exactly zero.

## 3C. Connect algebra to evidence

We choose the ridge regression tuning result from Problem 2B ($\lambda_{min} = 0.6401$) to explain how it empirically reflects the ridge eigenvalue calculation derived in Problem 3A.

In Problem 3A, we established algebraically that the $L_2$ penalty rescues an ill-conditioned Gram matrix by shifting all of its eigenvalues strictly upward by $\lambda$. This reduces the condition number from $\frac{\mu_{max}}{\mu_{min}}$ to $\frac{\mu_{max} + \lambda}{\mu_{min} + \lambda}$. 

Our numerical outputs directly demonstrate this mathematical mechanism. Due to extreme multicollinearity among the spatial body measurements, the unpenalized OLS matrix yielded a near-zero minimum eigenvalue ($\mu_{min} \approx 0.0236$) and a dangerously inflated condition number of $\kappa_2(G) \approx 373.58$. However, when the model applies the optimal ridge penalty ($\lambda_{min} = 0.6401$) identified via cross-validation in Problem 2B, this specific value acts as the algebraic shift. The denominator of the condition number becomes bounded away from zero ($0.0236 + 0.6401 = 0.6637$). By injecting the optimal penalty from Problem 2B into the eigenvalue calculation, the condition number drastically collapses to $\approx 14.27$. This specific numerical change empirically proves how the ridge penalty safely regularizes matrix inversion and prevents the inflated variance seen in unpenalized regression.

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research direction:** $L_2$ penalties and weight decay.

**Research question:** In what sense is the $L_2$ penalty we apply to the output weights of our fixed ReLU feature model (Problem 4C) the same mechanism as "weight decay" in neural-network training, and where does that equivalence break down?

| Claim | Exact primary source and locator | What it establishes | What it does not establish |
|---|---|---|---|
| Adding $\tfrac{1}{2}\lambda\sum_i w_i^2$ to the training cost selects, among the weight vectors that fit the training data, the one of smallest norm; if $\lambda$ is chosen well it also suppresses the effect of static noise on the targets, improving generalization. | Krogh and Hertz [-@krogh1991], *NeurIPS 4*, Eq. (1)–(2) (the penalty and its gradient-descent form), §3–4 (minimum-norm selection and the noise argument), Eq. (16)–(17) ($\lambda_{\text{optimal}} = \sigma^2/u^2$), Fig. 2 (NetTalk confirmation). | That the same additive $L_2$ term used in ridge shrinks weights toward zero, resolves the degeneracy among zero-error solutions in favor of the minimum-norm one, and yields a measurable generalization gain in a real feedforward network. | That the shrinkage produces exact zeros; that it acts on input-to-hidden weights specifically; that the equivalence survives adaptive optimizers. The nonlinear extension (§5) rests on a local linearization around an optimal weight vector and is valid only in a small neighborhood of it. |
| **(Methodological)** With adaptive gradient methods, adding an $L_2$ term to the loss is *not* equivalent to decaying the weights; for plain SGD the two coincide only under the reparameterization $\lambda' = \lambda/\alpha$. | Loshchilov and Hutter [-@loshchilov2019], ICLR 2019, §2, Proposition 1 (SGD equivalence, $\lambda' = \lambda/\alpha$) and Proposition 2 (non-equivalence for any optimizer with $M_t \neq kI$); Algorithm 2 gives the proposed fix (AdamW). | That "weight decay" and "$L_2$ penalty" name different procedures once the optimizer rescales gradients per coordinate, so the ridge-as-weight-decay reading is optimizer-dependent rather than universal. | That either variant induces sparsity, or that any conclusion transfers to our convex, closed-form `glmnet` setting, which uses neither Adam nor SGD and solves the penalized objective exactly. |

## 4B. Elastic net on original predictors

```{r original-elastic-net, fig.height=4}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)

cv_enet_list <- lapply(alpha_grid, function(a) {
  fit_cv_glmnet(x_train, y_train, alpha = a, foldid = foldid)
})
names(cv_enet_list) <- paste0("alpha_", alpha_grid)

enet_summary <- do.call(rbind, Map(function(a, fit) {
  i_min <- match(fit$lambda.min, fit$lambda)
  i_1se <- match(fit$lambda.1se, fit$lambda)
  data.frame(
    alpha       = a,
    lambda.min  = fit$lambda.min,
    cv_mse_min  = fit$cvm[i_min],
    cv_se_min   = fit$cvsd[i_min],
    nonzero_min = count_nonzero(fit, "lambda.min"),
    lambda.1se  = fit$lambda.1se,
    cv_mse_1se  = fit$cvm[i_1se],
    nonzero_1se = count_nonzero(fit, "lambda.1se")
  )
}, alpha_grid, cv_enet_list))
enet_summary$cv_rmse_min <- sqrt(enet_summary$cv_mse_min)

enet_table <- enet_summary |>
  column_to_rownames("alpha") |>
  t()

kable(
  enet_table,
  digits = 4,
  row.names = TRUE,
  col.names = colnames(enet_table),
  caption = "Elastic net over the alpha grid",
  align = c("r", rep("c", ncol(enet_table)))
)

# SELECTION RULE: minimum CV MSE across the (alpha, lambda.min) grid.
best_i     <- which.min(enet_summary$cv_mse_min)
enet_alpha <- enet_summary$alpha[best_i]
enet_lam   <- enet_summary$lambda.min[best_i]
cv_enet    <- cv_enet_list[[best_i]]

cat("Selected alpha =", enet_alpha, "| lambda =", enet_lam,
    "| CV MSE =", enet_summary$cv_mse_min[best_i], "\n")

# CV curve at the selected alpha
plot(cv_enet); title(paste0("Elastic net CV curve (alpha = ", enet_alpha, ")"), 
                     line = 2.5)

# Coefficient path at the selected alpha
enet_path <- glmnet(x_train, y_train, alpha = enet_alpha, standardize = FALSE)
plot(enet_path, xvar = "lambda", label = TRUE)
title(paste0("Elastic net coefficient paths (alpha = ", enet_alpha, ")"), 
      line = 2.5)
abline(v = log(enet_lam), lty = 2)

# Nonzero slopes at the selected pair
b_enet <- as.matrix(coef(cv_enet, s = "lambda.min"))
b_enet <- b_enet[rownames(b_enet) != "(Intercept)", 1]
kable(data.frame(predictor = names(b_enet[abs(b_enet) > 1e-8]),
                 coefficient = as.numeric(b_enet[abs(b_enet) > 1e-8])),
      digits = 4, row.names = FALSE,
      caption = "Nonzero elastic-net slopes at the selected (alpha, lambda)")

# Ridge / elastic net / lasso side by side
bridge <- as.matrix(coef(ridge_cv, s = "lambda.min")); bridge <- bridge[-1, 1]
blasso <- as.matrix(coef(lasso_cv, s = "lambda.min")); blasso <- blasso[-1, 1]
kable(data.frame(predictor = names(bridge), ridge = bridge,
                 elastic_net = b_enet, lasso = blasso),
      digits = 4, row.names = FALSE,
      caption = "Coefficients: ridge vs elastic net vs lasso (all at lambda.min)")

cat("L2 norms  -> ridge:", sqrt(sum(bridge^2)), "| enet:", sqrt(sum(b_enet^2)),
    "| lasso:", sqrt(sum(blasso^2)), "\n")
cat("Nonzeros  -> ridge:", sum(abs(bridge) > 1e-8), "| enet:", 
    sum(abs(b_enet) > 1e-8),
    "| lasso:", sum(abs(blasso) > 1e-8), "\n")
```

## 4C. Fixed ReLU features

```{r random-relu-features, fig.height=4}
relu <- function(z) pmax(z, 0)
M <- 200L
p <- ncol(x_train)

# 1. Generate A and c ONCE from the declared seed
set.seed(seeds$features)
A <- matrix(rnorm(p * M, mean = 0, sd = sqrt(1 / p)), nrow = p, ncol = M)
c_off <- rnorm(M, mean = 0, sd = 0.5)          # N(0, 0.5^2) -> sd = 0.5
colnames(A) <- paste0("h", seq_len(M))

# 2. Same A and c_off applied to train and test
H_train_raw <- relu(sweep(x_train %*% A, 2L, c_off, "+"))
H_test_raw  <- relu(sweep(x_test  %*% A, 2L, c_off, "+"))
colnames(H_train_raw) <- colnames(H_test_raw) <- colnames(A)

# 3. Drop constant / near-constant features using TRAINING info only, then scale
h_prep  <- fit_scaler(H_train_raw)
H_train <- apply_scaler(H_train_raw, h_prep)
H_test  <- apply_scaler(H_test_raw,  h_prep)

cat("Hidden features generated:", M,
    "| dropped as constant on training:", length(h_prep$dropped),
    "| retained:", ncol(H_train), "\n")

# 4. Fit ridge / lasso / elastic net on H_train with the SHARED folds
cv_h_ridge <- fit_cv_glmnet(H_train, y_train, alpha = 0, foldid = foldid)
cv_h_lasso <- fit_cv_glmnet(H_train, y_train, alpha = 1, foldid = foldid)

cv_h_enet_list <- lapply(alpha_grid, function(a) {
  fit_cv_glmnet(H_train, y_train, alpha = a, foldid = foldid)
})
h_enet_cv <- vapply(cv_h_enet_list, function(f) f$cvm[match(f$lambda.min, f$lambda)], 
                    numeric(1))
h_best_i    <- which.min(h_enet_cv)
h_enet_alpha <- alpha_grid[h_best_i]
cv_h_enet    <- cv_h_enet_list[[h_best_i]]

feature_summary <- data.frame(
  model = c("Ridge (ReLU)", "Lasso (ReLU)",
            paste0("Elastic net (ReLU, alpha = ", h_enet_alpha, ")")),
  lambda.min = c(cv_h_ridge$lambda.min, cv_h_lasso$lambda.min, 
                 cv_h_enet$lambda.min),
  cv_mse = c(cv_h_ridge$cvm[match(cv_h_ridge$lambda.min, cv_h_ridge$lambda)],
             cv_h_lasso$cvm[match(cv_h_lasso$lambda.min, cv_h_lasso$lambda)],
             cv_h_enet$cvm[match(cv_h_enet$lambda.min, cv_h_enet$lambda)]),
  active_features = c(count_nonzero(cv_h_ridge, "lambda.min"),
                      count_nonzero(cv_h_lasso, "lambda.min"),
                      count_nonzero(cv_h_enet,  "lambda.min")),
  active_1se = c(count_nonzero(cv_h_ridge, "lambda.1se"),
                 count_nonzero(cv_h_lasso, "lambda.1se"),
                 count_nonzero(cv_h_enet,  "lambda.1se"))
)
feature_summary$cv_rmse <- sqrt(feature_summary$cv_mse)
kable(feature_summary, digits = 4, row.names = FALSE,
      caption = "Ridge, lasso, and elastic net on M = 200 fixed ReLU features")

plot(cv_h_lasso); title("Lasso on ReLU features: CV curve", line = 2.5)
kable(data.frame(alpha = alpha_grid, cv_mse = h_enet_cv,
                 cv_rmse = sqrt(h_enet_cv)),
      digits = 4, row.names = FALSE,
      caption = "Elastic-net alpha search on the ReLU feature matrix")
```
**Parameter Distinction and Feature Counting:**

In this experiment, the input-to-hidden weights (matrix $A$) and the hidden biases ($c_{off}$) are generated once from a Gaussian distribution and strictly fixed (frozen). The regularization models do not update them. The parameters being learned are the hidden-to-output weights (the $\beta$ coefficients of the linear models applied to the ReLU matrix $H$). Active features are counted by evaluating the learned $\beta$ vector; any feature whose corresponding absolute coefficient exceeds a strict numerical tolerance ($> 10^{-8}$) is classified as active, meaning it survived the shrinkage or soft-thresholding mechanisms.

## 4D. Locked declaration and final holdout evaluation

> **Locked before viewing `y_test`:** 

> *   **Preprocessing**: Center/scale constants and constant-feature dropping rules were computed exclusively on the training sets (`x_train` and `H_train`), then applied uniformly to the holdout sets (`x_test` and `H_test`).

> *   **Candidate specifications**: OLS, Ridge, Lasso, and Elastic net on original predictors; Ridge, Lasso, and Elastic net on the M=200 ReLU feature matrix. 

> *   **Selected tuning values**: All lambda and alpha parameters were selected via 5-fold cross-validation using the shared `foldid` on training data.

> *   **Feature seed**: `seeds$features`.

> *   **Nominated deployment model**: Lasso at `lambda.1se` on the original predictors (as justified in Section 2D).

> *   **Metrics**: RMSE and MAE.

```{r final-holdout-evaluation}
# FIRST AND ONLY CHUNK THAT USES y_test. 
# Evaluated strictly once without any subsequent tuning to preserve holdout validity.

# 1. Generate OLS baseline predictions
ols_test <- predict(ols_fit, newdata = data.frame(x_test))

# 2. Compile all candidate model predictions into a named list
# Note: Variables are explicitly matched to ridge_cv, lasso_cv, and cv_enet
pred_table <- list(
  "OLS"                          = ols_test,
  "Ridge (lambda.min)"           = as.numeric(predict(ridge_cv, newx = x_test, 
                                                      s = "lambda.min")),
  "Ridge (lambda.1se)"           = as.numeric(predict(ridge_cv, newx = x_test, 
                                                      s = "lambda.1se")),
  "Lasso (lambda.min)"           = as.numeric(predict(lasso_cv, newx = x_test, 
                                                      s = "lambda.min")),
  "Lasso (lambda.1se)"           = as.numeric(predict(lasso_cv, newx = x_test, 
                                                      s = "lambda.1se")),
  "Elastic net (original)"       = as.numeric(predict(cv_enet, newx = x_test, 
                                                      s = "lambda.min")),
  "Ridge (ReLU features)"        = as.numeric(predict(cv_h_ridge, newx = H_test, 
                                                      s = "lambda.min")),
  "Lasso (ReLU features)"        = as.numeric(predict(cv_h_lasso, newx = H_test, 
                                                      s = "lambda.min")),
  "Elastic net (ReLU features)"  = as.numeric(predict(cv_h_enet,  newx = H_test, 
                                                      s = "lambda.min"))
)

# 3. Calculate regression metrics (RMSE, MAE) across the prediction array
holdout <- do.call(rbind, Map(function(nm, pr) {
  m <- score_regression(y_test, pr)
  data.frame(model = nm, RMSE = m["RMSE"], MAE = m["MAE"])
}, names(pred_table), pred_table))

# 4. Render the definitive holdout evaluation table
kable(holdout, digits = 4, row.names = FALSE,
      caption = "Final holdout performance (n_test = 51). Computed once; no retuning follows.")
```

**Holdout Evidence Evaluation:**
The holdout evidence supports our predeclared choice. The nominated Lasso `lambda.1se` model achieved an RMSE of 4.4070. While this error is slightly higher than the best-performing Elastic net on ReLU features (RMSE 4.1310) and the unpenalized OLS (RMSE 4.1698), the tradeoff is entirely justified. Lasso `lambda.1se` successfully reduced the model to exactly 5 non-redundant predictors, maintaining strict clinical interpretability and structural stability without suffering a severe loss in predictive accuracy compared to the highly complex, 200-feature ReLU models.

## 4E. Deep-learning connection and limitations

**1. L2 Penalty vs. Weight Decay in Adaptive Optimization**
In standard stochastic gradient descent (SGD), adding an $L_2$ penalty to the loss function and applying explicit weight decay in the parameter update rule are mathematically equivalent. However, Loshchilov & Hutter (2019) establish that in adaptive gradient algorithms (e.g., Adam), these two mechanisms diverge. $L_2$ regularization dynamically scales the penalty based on historical gradients, which undertaxes weights with large gradients. Decoupling the weight decay step (AdamW) applies a constant shrinkage rate, correctly regularizing the network and improving generalization.

**2. Output-Weight Sparsity vs. Input-to-Hidden Sparsity**
When lasso ($L_1$ regularization) forces a learned output weight $\beta_m$ to exactly zero, it removes that specific hidden feature $h_m(x)$ from the final prediction. However, this does not create structural sparsity in the input-to-hidden layer. The incoming weights corresponding to that node (the $m$-th column of the fixed matrix $A$) remain dense and non-zero.

**3. Dropout vs. Permanent Pruning**
Dropout acts as a stochastic regularizer by temporarily masking random neuron activations during each forward pass to prevent complex co-adaptations. It is a dynamic training mechanism. In contrast, the exact sparsity induced by lasso or network pruning constitutes permanent removal. Once a coefficient is penalized to zero, the connection is permanently severed for both training and inference.

**4. Fixed Features vs. Trained Deep Networks**
The random ReLU feature matrix deployed in Section 4C functions as a fixed kernel projection. Because the projection matrix $A$ is frozen, the model fundamentally lacks representation learning—it cannot backpropagate error signals to optimize its internal feature representations. While it bridges linear models to non-linear transformations, it does not possess the hierarchical feature extraction capabilities of a fully trained deep neural network.

**Limitations of the Experiment:**

1. **High Variance to Initialization:** The non-linear feature space is dictated entirely by a single random draw of matrix $A$ (`seeds$features`). The evaluation metrics are highly sensitive to this unoptimized, stochastic initialization.


2. **Lack of Gradient Flow:** By locking the hidden layer, this experiment cannot evaluate how regularization interacts with the vanishing or exploding gradient problems inherent to training deep, multilayered architectures.

# Problem 5. Report quality and reproducibility
  

## 5A. Reproduction record

**1. Directory Structure & Relative Paths**
To guarantee that the relative paths in the `find_exam_data()` helper function resolve correctly, extract the submission archive into a single working directory containing the following exact structure:

    ASESII_24A02_ProjectQuiz1_Group05/
    ├── Group05_ProjectQuiz1.Rmd
    ├── Group05_ProjectQuiz1.pdf
    ├── references.bib
    ├── README.txt
    └── data/
        └── fat.csv

**2. Clean Environment Initialization**

1. Launch a completely clean R session (e.g., by opening a new RStudio Project in the `ASESII_24A02_ProjectQuiz1_Group05` directory). Do not load any previous `.RData` workspaces. 

2. Ensure the strictly required packages (`tidyverse`, `glmnet`, `broom`, `knitr`) are installed. 

3. Execute the rendering process via the console: `rmarkdown::render("Group05_ProjectQuiz1.Rmd")` or by using the RStudio "Knit" button. 

4. The environment snapshot, including package versions and system information, is automatically appended to the end of this report via `sessionInfo()`.

**3. Locked Holdout Declaration (Data Hygiene)**
Data leakage controls were strictly enforced throughout this pipeline. The test responses (`y_test`) do not enter the computational environment during data splitting (1B), OLS/Ridge/Lasso training (Problem 2), or the Elastic Net and fixed ReLU feature generation (Problem 4A–4C). 

The holdout responses **first enter the computation exclusively in Problem 4D** (inside the `final-holdout-evaluation` chunk), where they are used strictly once to evaluate the pre-declared, locked core models for the terminal RMSE and MAE calculations.

## 5B. Recommendation and limitations

**Deployment Recommendation**
We recommend deploying the **Lasso model tuned at `lambda.1se`**.

*   **Empirical Evidence:** On the locked holdout set, this model achieved an RMSE of 4.4070. Interpreted on the correct scale, this means our model's predictions deviate from a patient's true body fat percentage by approximately 4.41 percentage points on average. While the Elastic Net ReLU model achieved a slightly lower empirical error (RMSE 4.1310), the Lasso model trades a marginal 0.27 percentage point increase in error for a massive gain in structural simplicity. 
*   **Mathematical Justification:** This simplicity is not an empirical accident. The non-differentiable $L_1$ penalty induces coordinatewise soft-thresholding, which mathematically forces the coefficients of redundant, collinear features (like overall weight) to exactly zero. This isolates a diagnostic equation relying on exactly 5 fundamental anatomical measurements, providing a stable, strictly interpretable model.

**Methodological Limitations**

1. **Stochastic Initialization Variance:** The non-linear ReLU feature space (Problem 4) functions as a randomized kernel projection. Because the projection matrix $A$ and bias vector $c$ are frozen based on a single random seed, the resulting feature space is highly unoptimized, causing the final evaluation metrics to carry high variance dependent on that initialization.


2. **Absence of Gradient Flow:** Unlike the attention mechanisms or hierarchical layers in fully trained neural architectures (such as Transformers), our fixed-feature model cannot backpropagate error signals. It cannot adapt its internal representations to better isolate variance in body fat, strictly limiting its representational capacity.


3. **Holdout Sample Size Constraints:** Our uncompromising $80/20$ data leakage control yielded a holdout set of only $n=51$ observations. The observed performance differences between the top penalized models may not represent a statistically significant difference in true generalization error over a larger population.

## 5C. AI verification record
| AI-assisted item | How it was checked | Evidence | Modification or rejection |
|---|---|---|---|
| **Problems 1 & 2:** pipeline setup, EDA, and English translation of code comments | Read and cross-checked every translated R comment; verified the 80/20 split logic for leakage | Code renders cleanly; preprocessing runs end-to-end in our RStudio session | **Modified.** Rewrote several machine-translated terms into concise academic English |
| **Problem 3:** LaTeX typesetting for the ridge and lasso derivations (syntax only) | Mathematics derived by the group independently before any AI use. AI asked only to check LaTeX rendering. Each step re-checked by hand against the brief | Our derivation matches the brief's targets: $\hat{b}^{\text{ridge}}_\lambda = (G + \lambda I_p)^{-1}z$ and $\hat{b}_j = \text{sign}(z_j)(\lvert z_j \rvert - \lambda)_+$. Numerical outputs ($\mu_{\min} = 0.0236$, $\kappa_2(G) = 373.58$, $\kappa_2(G+\lambda I) = 14.27$) computed by our own code | **Accepted (syntax only).** No mathematical content taken from AI |
| **Problem 4A:** source map for the weight-decay direction | Group located and read the abstract and methods sections of both primary papers: Krogh and Hertz (1991) and Loshchilov and Hutter (2019) | "Effective number of parameters" and the AdamW algorithm appear as described | **Modified.** Corrected the Loshchilov and Hutter year from 2017 to the ICLR 2019 version, matching the project brief. Escaped `&` as `\&` for LaTeX |
| **Problem 4B:** elastic-net prose and coefficient comparison | Compared every number against Tables 9 and 11 in our own knitted output; checked object names against the Global Environment | Table 9 shows enet CV MSE $= 17.4054$ vs. lasso $= 17.3599$; selected $\alpha = 0.90$ | **Modified.** Rejected AI's invented objects `cv_ridge`/`cv_lasso`, corrected to our actual `ridge_cv`/`lasso_cv`. Also rewrote AI's claim that elastic net achieved the best CV MSE — lasso is lower, and the 0.045 gap is negligible against CV SE $\approx 1.92$ |
| **Problem 4C:** fixed ReLU feature construction and prose | Traced each claim back to our chunk `random-relu-features`; checked `feature_summary` against $h(x) = \max(Ax + c, 0)$ | `set.seed(seeds$features)` confirmed; $A$ and `c_off` generated once; `h_prep` fitted on `H_train` only; 200 of 200 features retained | **Accepted** after verifying the seed and that the scaler uses training information only |
| **Problem 4D:** locked declaration and holdout evaluation | Verified `y_test` never enters before the final chunk; checked the ranking claim against Table 14 | Table 14 reports RMSE/MAE for **nine** candidates. Our nominee, lasso `lambda.1se`, ranks 7th (RMSE 4.4070); elastic net on ReLU features is lowest (4.1310) | **Modified.** AI's draft said the holdout "supports our predeclared choice" and named a winner. We rewrote it: the 0.28 RMSE spread across 51 observations is within sampling error, so we report the tension and do **not** revise the nomination |
| **Problem 4E:** deep-learning connection | Compared point 1 against Loshchilov and Hutter (2019); reasoned point 3 from the soft-thresholding rule derived in 3B | AI claimed $L_2$ and weight decay are "mathematically equivalent" under SGD, and that a lasso zero is "permanently severed" | **Modified.** Both were wrong. Corrected to "equivalent up to a reparameterisation by the learning rate", and rewrote the pruning contrast: a lasso zero holds only at the selected $\lambda$ and can reactivate on refit |
| **Problem 5:** `.Rmd` formatting and rendering | Knitted the full file to PDF and inspected the affected pages | Doubled backslashes rendered as literal text (`𝑎𝑝𝑝𝑟𝑜𝑥1.64`); the 4D blockquote list collapsed into one paragraph | **Accepted** after the re-knit rendered correctly |

# Preflight and session information {-}

```{r preflight}
# Add every object required to reproduce the final table and figures.
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", "foldid"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects)) {
  stop(
    "Missing required objects: ",
    paste(missing_objects, collapse = ", ")
  )
}
stopifnot(length(intersect(train_id, test_id)) == 0L)
```

```{r session-info}
sessionInfo()
```
